Solving an Exam Scheduling Problem Using a Genetic Algorithm

### 2009-12-01 trunk/Paper/paper.tex
Updated table in paper, not perfect, but good enough for the moment.
Paper update, 3D graphs in it

\documentclass[letterpaper]{article}
\usepackage{aaai}
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\usepackage{url}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PDFMARK for TeX and GhostScript
% Uncomment and complete the following for metadata if
% your paper is typeset using TeX and GhostScript (e.g
% if you use .ps or .eps files in your paper):
% \special{! /pdfmark where
% {pop} {userdict /pdfmark /cleartomark load put} ifelse
% [ /Author (John Doe, Jane Doe)
% /Title (Paper Title)
% /Keywords (AAAI, artificial intelligence)
% /DOCINFO pdfmark}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PDFINFO for PDFTeX
% Uncomment and complete the following for metadata if
% your paper is typeset using PDFTeX
\pdfinfo{
  /Title (Solving an Exam Scheduling Problem Using a Genetic Algorithm)
  /Subject (Input The Proceedings Title Here)
  /Author (Dave, Kordalewski;
           Caigu, Liu;
           Kevin, Salvesen;)
} 
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Uncomment only if you need to use section numbers
% and change the 0 to a 1 or 2
% \setcounter{secnumdepth}{0}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

  \subsection{Genetic Algorithm Search}
    Although there are many readily available open source genetic algorithm software
    packages available (see JGAP \cite{website:JGAP}, Jenes\cite{website:jenes}) we 
    decided to write a simple, yet general, genetic algorithm package, in Java, that
    can interface with data types designed by others, so long as they implement a few
    necessary methods. The package we wrote is more than adequate for investigating
    the scheduling problem under consideration here.

    \subsubsection{General Description}
      The genetic algorithm maintains a population of schedules and constructs subsequent
      generations by adding new random schedules, making copies of high quality schedules 
      in the present generation, applies the mutation operator to schedules, or uses the
      crossover operator on 2 schedules. When performing mutation or crossover, the 
      schedules are selected with a probability proportional to their fitness, allowing 
      the schedules with high quality tobe selected more frequently and make more 
      "offspring".

      The free parameters in our GA are the size of the population, number of
      generations to simulate, and proportions of new generations toi be made by the copy, 
      random, mutation, and crossover methods.

@@

  \subsection{Options for the Genetic Algorithm}
    \subsubsection{Scheduling Problem Instances}
      We generated 2 sets of 3 scheduling problem instances to test the algorithms above with. 
      The package ScheduleInstanceGenerator in our code repository manages the generation of 
      random problem instances with many free parameters to specify characteristics of the 
      output instance.

@@

      \begin{tabular}{|c|c|c|c|c|c|c|}
      \hline  & Tiny & Medium & Large & Tiny /w Cor. & Med /w Cor. & Large /w Cor. \\ 
      \hline d & 7 & 10 & 7 & 7 & 10 & 7 \\ 
      \hline t & 5 & 8 & 8 & 5 & 8 & 8 \\ 
      \hline $|C|$ & 20 & 200 & 603 & 20 & 200 & 600 \\ 
      \hline $|R|$ & 4 & 10 & 43 & 4 & 10 & 43 \\ 
      \hline $|S|$ & 50 & 300 & 21945 & 50 & 300 & 21945 \\ 
      \hline $|L_s|$ & 1-5 & 1-6 & 1-5 & 2-6 & 2-6 & 2-6 \\ 
      \hline \# Course Types & N/A & N/A & N/A & 4 & 10 & 20 \\ 
      \hline 
	\end{tabular}

	  The full input files for these instances are available in our repository in the 
	  directory "Instance Input Files".

    \subsubsection{Selecting Parameters for the GA}
      [\ldots]
  \subsection{Comparison Between the Different Searches}
    [\ldots]

\section{Evaluation}
  \subsection{Success}
    [\ldots]
    Seeing as our fitness function gives a very high penalty for schedules that have direct
    conflicts (either a student with two exams at the same moment, or two different exams
    in a same room at the same moment), if such a schedule exists given the input data, our
    algorithms will very quickly tend to reject all such schedules.

\section{Conclusion}
  We come up with a conclusion that in this particular problem, genetic algorithm
  is better than random search, and it is better than hill-climbing if the input 
  is a large data set. But it may not be the best solution for this problem, since
  the Mutate Search performs way better.

\bibliographystyle{aaai}
\bibliography{paperBib}
\end{document}

>>>>

\documentclass[letterpaper]{article}
\usepackage{aaai}
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\usepackage{url}
\usepackage[table]{xcolor}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{float}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PDFMARK for TeX and GhostScript
% Uncomment and complete the following for metadata if
% your paper is typeset using TeX and GhostScript (e.g
% if you use .ps or .eps files in your paper):
% \special{! /pdfmark where
% {pop} {userdict /pdfmark /cleartomark load put} ifelse
% [ /Author (John Doe, Jane Doe)
% /Title (Paper Title)
% /Keywords (AAAI, artificial intelligence)
% /DOCINFO pdfmark}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PDFINFO for PDFTeX
% Uncomment and complete the following for metadata if
% your paper is typeset using PDFTeX
\pdfinfo{
  /Title (Solving an Exam Scheduling Problem Using a Genetic Algorithm)
  /Subject (Input The Proceedings Title Here)
  /Author (Dave, Kordalewski;
           Caigu, Liu;
           Kevin, Salvesen;)
} 
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Uncomment only if you need to use section numbers
% and change the 0 to a 1 or 2
% \setcounter{secnumdepth}{0}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

  \subsection{Genetic Algorithm Search}
    Although there are many readily available open source genetic algorithm software
    packages available (see JGAP, Jenes) we 
    decided to write a simple, yet general, genetic algorithm package, in Java, that
    can interface with data types designed by others, so long as they implement a few
    necessary methods. The package we wrote is more than adequate for investigating
    the scheduling problem under consideration here.

    \subsubsection{General Description}
      The genetic algorithm maintains a population of schedules and constructs subsequent
      generations by adding new random schedules, making copies of high quality schedules 
      in the present generation, applies the mutation operator to schedules, or uses the
      crossover operator on 2 schedules. When performing mutation or crossover, the 
      schedules are selected with a probability proportional to their fitness, allowing 
      the schedules with high quality to be selected more frequently and make more 
      "offspring".

      The free parameters in our GA are the size of the population, number of
      generations to simulate, and proportions of new generations to be made by the copy, 
      random, mutation, and crossover methods.

@@

  \subsection{Options for the Genetic Algorithm}
    \subsubsection{Scheduling Problem Instances}
      We generated 2 sets of 3 scheduling problem instances to test the algorithms above with. 
      The package \texttt{ScheduleInstanceGenerator} in our code repository manages the generation of 
      random problem instances with many free parameters to specify characteristics of the 
      output instance.

      The first set of instances were generated with the parameters in the table below. 
      The rooms and courses and students were given arbitrary names and the courses that each 
      student attended were selected with uniform probability from the complete set of courses.

      The second set was generated with a sort of correlation between the courses that each student 
      took, rendering the dataset more true to the sort of data that would be found in real-world 
      instances of the exam scheduling problem. The set of courses that each student takes was 
      selected in the following way. Each student was assigned, with uniform probability, a \emph{major} 
      and a random number of courses between 2 and 6. Each course was then selected, with probability 
      0.7, from among the $|S|/\#Majors$ courses in their major, or, with probability 0.3, uniformly from 
      the set of all courses. We can expect that each student takes 70\% of their courses from a small 
      subset of the courses, representing the field that they specialize in, and 30\% from other fields. 
      This simulates the way in which we expect student bodies to choose their courses, though the true 
      data is likely to exhibit far more structure, due to program requirements, notorious professors, etc.

      The 3 sizes vary from manageable to large. The smallest could potentially be solved exactly, 
      with methods not under consideration here, while the others have search spaces much too large 
      for any hope of an exact solution. For the largest, which simulates a UofT exam schedule, a 
      single evaluation of the fitness function took 0.2 seconds on the fastest machine owned by any 
      of the authors.

@@

      \definecolor{light-gray}{gray}{0.87}
      \rowcolors{1}{light-gray}{white}
      \begin{center}
      \begin{small}
      \begin{tabular}{*{7}{c}}
      \hline 
      	& S 	& M 	& L 	& $S_{Cor}$ & $M_{Cor}$ & $L_{Cor}$ \\ 
      \hline \hline
       d 		& 7 	& 10	& 7 	& 7 	& 10 	& 7 	\\ 
       t 		& 5		& 8 	& 8 	& 5 	& 8 	& 8 	\\ 
       $|C|$	& 20 	& 200 	& 603 	& 20 	& 200 	& 600 	\\ 
       $|R|$	& 4 	& 10 	& 43 	& 4 	& 10 	& 43	\\ 
       $|S|$	& 50 	& 300 	& 21945 & 50 	& 300 	& 21945	\\ 
       $|L_s|$	& 1-5 	& 1-6 	& 1-5 	& 2-6 	& 2-6 	& 2-6 	\\ 
       \begin{tiny}$\#$Maj\end{tiny}	& N/A 	& N/A 	& N/A 	& 4 	& 10 	& 20 	\\ 
       \textsc{sss}	& $10^{182}$ & $10^{1840}$ & $10^{6695}$ & $10^{182}$ & $10^{1840}$ & $10^{6689}$ \\
      \hline 
	  \end{tabular} 
	  \end{small}
	  \end{center}

	  The large data set was created with parameters inspired from the University of 
	  Toronto's Faculty of Arts and Science.
	  The full input files for these instances are available in our repository in the 
	  directory \emph{Instance Input Files}.

    \subsubsection{Selecting Parameters for the GA}
      In order to select parameters for the genetic algorithm (that is, the population, 
      number of generations to simulate, proportions of copies, randoms, mutations, and 
      crossovers for the next generation) we performed tests using the small uncorrelated 
      problem instance.

      \paragraph{Generation Parameters}
        We undertook a search of the space of parameters for forming new generations.

        In the graph below, each data point represents the average of 5 trials. To make 
        the scale clearer, the colour on the surface corresponds to the fitness, where 
        the highest fitnesses are red. The number of copies is fixed at 1, and the number 
        of mutations (m) and crossovers (c) are shown on the horizontal axes. The total 
        size of the population is fixed, so the number of random individuals used at each 
        point is 60?-(m+c). This corresponds to the distance from the diagonal.

        \begin{figure}[H]
  		  \centering
  		  \includegraphics[width=8cm]{GAParametersData}
  	      \caption{Effectiveness of the GA's proportion parameters}
  	      \label{graph:GAParam1}
		\end{figure}

        The graph above indicates that for a fixed number of evaluations, best results are 
        achieved by selecting parameters that have the majority of individuals in the next 
        generation be generated by crossovers of the most fit schedules in the previous 
        generation. In order to achieve the very highest fitness, a few randoms or mutations 
        should also be incorporated.

        This test was done for only the smallest problem instance, and is infeasible for the 
        larger instances. We chose generation proportion parameters to approximate the optimal 
        values seen here, under the assumption that the larger problems exhibited this sort of 
        preference.

      \paragraph{Size Parameters}
        Blah blah blaj. We investigated the population size and

        \begin{figure}[H]
  		  \centering
  		  \includegraphics[width=8cm]{MEDIUM-GA_Size_Data}
  	      \caption{Effectiveness of the GA's population parameters}
  	      \label{graph:GAParam2}
		\end{figure}

  \subsection{Comparison Between the Different Searches}
    [\ldots]
    \begin{figure}[h!]
  	  \centering
  	  \setlength\fboxsep{0pt}
	  \setlength\fboxrule{0.5pt}
	  \fbox{\includegraphics[width=8.35cm]{HUGE-TYPE2_FullAverageStats}}
      \caption{Search comparison on a large data set}
	\end{figure}

\section{Evaluation}
  \subsection{Success}
    [\ldots]
    Seeing as our fitness function gives a very high penalty for schedules that have direct
    conflicts (either a student with two exams at the same moment, or two different exams
    in a same room at the same moment), if such schedules exists given the input data, 
    our algorithms will tend to very quickly reject them.

\section{Conclusion}
  [\ldots]
  We come up with a conclusion that in this particular problem, genetic algorithm
  is better than random search, and it is better than hill-climbing if the input 
  is a large data set. But it may not be the best solution for this problem, since
  the Mutate Search performs way better.

  \nocite{*}
\bibliographystyle{aaai}
\bibliography{paperBib}
\end{document}
